Results of two stellar occultations by dwarf planet (1) Ceres

### 2017-08-16 Ceres/Old_versions/ArtigoV6/CeresV6.tex
mod 2017 ago16

\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail:
altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6}$,
M. Assafin$^{1}$,\newauthor
R. Vieira-Martins$^{1,5}$
T. Blank$^{4}$, 
J. Brooks$^{4}$, 
J. Broughton$^{7}$, 
R. F. Dantowitz$^{8}$, \newauthor
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$, 
T. George$^{4}$, 
P. D. Maley$^{4,9}$, \newauthor
B. Timerson$^{4}$, 
N. S. Weber$^{8}$\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
$^{8}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{9}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA}

\begin{document}

@@

\begin{abstract}
We report the results from two multi-chord stellar occultations by (1) Ceres that occurred in 2010 August 17, observed at Brazil, and 2013 October 25, observed at USA. Four positive chords were obtained for the 2010 occultation and nine for the 2013 occultation. An elliptic model was used to fit the apparent limb of Ceres. Two limb fitting solution was studied. The first one is a nominal solution with an indeterminate polar aspect angle. The second one, we constrain the pole coordinates as given by \cite{Drummond2014}. Assuming a Maclaurin spheroid, we determine an equatorial radius of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$.
\end{abstract}

@@

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of the kilometre, in a time when the uncertainties were often ten times larger. The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed. Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}. The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work. The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords. Throughout the paper we shall refer to these events as the `2010' and the `2013 occultation'.

@@

This work is organized as follows. In Sections 2 and 3 we analyse, the 2010 and the 2013 events, respectively. The comparison of both results to those on the literature is carried out in Section 4.

\section{The 2010 occultation}

@@

%Asteroidal occultations of bright stars are routinely predicted by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.}, on behalf of the International Occultation Timing Association. Within this context it was identified that on 17 August 2010 Ceres would occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $\text{K}=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
%
On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRS position is \citep{Zacharias2004}:
% UCAC4 313-111823
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
    \delta = -27\degr 26\arcmin 38\arcsec.890.
  \end{array}
\right.
\label{Eq: 2010-coord}
\end{equation}
%
% Abaixo a coordenada UCAC4 aplicada de Movimento Próprio
%
%\begin{equation}
%\left\{ 
%  \begin{array}{l l}
%    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.009\\
%    \delta = -27\degr 26\arcmin 38\arcsec.866.
%  \end{array}
%\right.
%\label{Eq: 2010-coord}
%\end{equation}
%
The shadow's path would cross South America, the Atlantic and reach the south part of Africa.

@@

Observations were carried out in Brazil at five stations as displayed in Table \ref{Tab: obs-2010} and Fig. \ref{Fig: Ceres-2010-map}. The occultation was detected on four of them. UFSC had a negative chord. From the positive ones, INPE started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope & Exposure & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & Camera    &        & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
            & 825 m                &       &     & 22:40:20 $\pm$ 5 &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34'45.1'' W &       & 1 s & Positive & J. I. B. Camargo \\
 OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
            & 1864 m               &       &     & 22:41:55.3 $\pm$ 0.7 &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start: 22:39:44 & T. Maldonado\\
 INPE      & 975 m                & SBIG ST7 &     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
            & 910 m                & SBIG STL6E &     & 22:39:56 $\pm$ 13 & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31'20.5'' W &       & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  & A. J. T. Mello\\
            & 20 m                 &       &  &         & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}. The light curves were normalized to an unocculted star flux. Additionally, with exception of INPE, they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

@@

The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}. The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

@@

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star. The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.

@@

The apparent oblateness is related to the true one $\epsilon$ through
%
\begin{equation}
\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
\label{Eq: TrueObla}
\end{equation}
%
where $\zeta$ is the polar aspect angle, encompassed by the (true) polar semiminor axis and the line of sight. Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid.

@@

Two possible solutions were considered for the limb fitting. The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large error bar on the position angle. Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.

From observations by adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha = (287 \pm 3) \degr, \quad %\\
\delta = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Eq: Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore Eq. \ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq. \ref{Eq: Pole}. We call this the ``pole-constrained solution''.

@@

Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$. Clearly, the coordinates of Ceres' pole (Eq. \ref{Eq: Pole}) can impose a strong constraint on the position angle, as the next solution shows.

@@

At the moment of the occultation, the coordinates (Eq. \ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$. Exploration of the parameter space, restricted to ellipses whose position angle lies on this range, results in the pole-constrained solution. The related physical parameters are displayed on the third column of Table \ref{Tab: results}, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Results of limb fitting to the data of the 2010 and 2013 events.\label{Tab: results}}
  \begin{tabular}{@{}lcccc}
  \hline
     Solution & 2010/Nominal & 2010/Pole-constrained & 2013/Nominal & 2013/Pole-constrained \\
\hline
Equatorial diameter (km) & 982 $\pm$ 14 & 972 $\pm$ 6  & 971 $\pm$ 7  & 971 $\pm$ 7\\
%Equatorial diameter (km) & 491 $\pm$ 7  & 486 $\pm$ 3  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
True oblateness        & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 77 $\pm$ 6    & 78 $\pm$ 6\\
$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 13 $\pm$ 16   & 13 $\pm$ 16\\
$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
\hline
\end{tabular}
\textbf{Notes.} Error bars are at 1-sigma level.
(*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
\end{minipage}
\end{table*}

@@

\section{The 2013 occultation}

@@

The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $K = 8.76$. Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is :
%
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 11^{h}57^{m}52^{s}.7641\\
    \delta = +09\degr 07\arcmin 49\arcsec.835
  \end{array}
\right.
\end{equation}
%
The occultation could only be visible in the United States, before dawn. The shadow would then cross the Atlantic and Africa, but during daylight time.

\begin{figure*}
\begin{centering}
\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013.png}}
\hspace{5mm}
\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.png}}
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA. Left panel: Overview of the occultation. The big red dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right. The blue dots are the sites that have observed the event. Right Panel: Upper view of the occultation over the sites that observed the event. Red points are the centre of the shadow separated by 5 seconds. \label{Fig: Ceres-2013-map}}
\end{centering}
\end{figure*}

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for the observing stations of the 2013 event.\label{Tab: obs-2013}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope: & Camera  & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio    &           & Egress    & \\          
\hline
 Hampton & 70$\degr$48'59.7'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$53'52.8'' N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 &   \\
            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 &   \\
 & & & & & \\
 Topsfield & 70$\degr$55'16.6'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$37'55.9'' N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 &   \\
            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 &   \\
 & & & & & \\
 Brookline & 71$\degr$08'14.5'' W & 64 cm & Infinity2-1R & Positive & N. Weber \\
  &42$\degr$18'27.4'' N & f/9.6 & 0.015 s     & Duration only &  R. Dantowitz  \\
            & 109 m     &       &     & (14.93 $\pm$ 0.01) s &   \\
 & & & & & \\
 Winchester & 78$\degr$14'39.6'' W & 36 cm & PC164C & Positive & J. Brooks \\
  &39$\degr$16'21.5'' N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 &   \\
            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 &   \\
 & & & & & \\
 Greenbelt & 76$\degr$52'09.4'' W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
  &38$\degr$59'12.1'' N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 & D. Dunham  \\
            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 &   \\
 & & & & & \\
 Alexandria & 77$\degr$02'28.3'' W & 7 cm & Watec120N & Positive & P. Maley \\
  &38$\degr$49'19.1'' N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 &   \\
            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 &   \\
 & & & & & \\
 Owings & 76$\degr$38'06.3'' W & 25 cm & PC164C  & Positive & C. Ellington \\
  &38$\degr$41'26.5'' N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 &   \\
            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 &   \\
 & & & & & \\
 Mechanicsville & 77$\degr$23'06.7'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$41'26.1'' N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 &   \\
            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 &   \\
 & & & & & \\
 Varina & 77$\degr$19'49.3'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$25'58.6'' N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 &   \\
            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 &   \\

@@

All videos were converted to FITS images and the photometry of the target was obtained via PRAIA \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of a reference star when such was available on the field.

>>>>

\title[Results of two stellar occultations by Ceres]{Results of two stellar occultations by dwarf planet (1) Ceres}
\author[A. V. Raveendran and A. N. Other]{A. V. Raveendran$^{1}$\thanks{E-mail:
email@address (AVR); otheremail@otheraddress (ANO)} and A. N.
Other$^{2}$\footnotemark[1]\thanks{}\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77,
CEP 20921-400 Rio de Janeiro - RJ, Brazil}
\begin{document}

@@

\begin{abstract}
Abstract
\end{abstract}

@@

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of the kilometre, in a time when the uncertainties were often ten times larger. The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four like this have been observed. Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}. The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work. The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords. Throughout the paper we shall refer to these events as the `2010' and the `2013 occultation'.

@@

This work is organized as follows. In Sections 2 and 3 we analyse, on this order, the 2010 and the 2013 events. The comparison of both results to those on the literature is carried out in Section 4.

\section[]{The 2010 occultation}

@@

%Asteroidal occultations of bright stars are routinely predicted by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.}, on behalf of the International Occultation Timing Association. Within this context it was identified that on 17 August 2010 Ceres would occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $\text{K}=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
%
On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
%
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
    \delta = -27\degr 26\arcmin 38\arcsec.890.
  \end{array}
\right.
\end{equation}
%
The shadow's path would cross South America, the Atlantic and reach the south part of Africa.

@@

Observations were carried out in Brazil at five stations as displayed in Table~1 and Fig.~\ref{Fig: Ceres-2010-map}. The occultation was detected on four of them. Among these, one started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only 3.9 km s$^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
     Site & Longitude & Telescope & Exposure & Result & Observer  \\
          & Latitude  & Aperture  & Cycle time & Ingress & \\          
          & Height    & Camera    &    S/N    & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
            & 825 m                &       &     & 22:40:20 $\pm$ 5 &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34'45.1'' W &       & 1 s & Positive & J. I. B. Camargo \\
 OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
            & 1864 m               &       &     & 22:41:55.3 $\pm$ 0.7 &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start: 22:39:44 & T. Maldonado\\
 INPE      & 975 m                & SBIG ST7 &     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
            & 910 m                & SBIG STL6E &     & 22:39:56 $\pm$ 13 & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31'20.5'' W &       & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  & A. J. T. Mello\\
            & 20 m                 &       &  &         & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}. The light curves were normalized to an unocculted star flux. Additionally --with exception of INPE-- they were were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

@@

The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}. The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the occultation light curves are shown in Fig.~\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

@@

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis --$a^\prime$ and $b^\prime$, respectively--, position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star\footnote{The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east.}. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.

@@

The apparent oblateness is related to the true one $\epsilon$ through
%
\begin{equation}
\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
\label{TrueObla}
\end{equation}
%
where $\zeta$ is the polar aspect angle, encompassed by the (true) polar semiminor axis and the line of sight. Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid.

@@

Two possible solutions were considered for the limb fitting. The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large error bar on the position angle. Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \eqref{TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.

From observations by adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha = (287 \pm 3) \degr, \quad %\\
\delta = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore \eqref{Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from \eqref{Pole}. We call this the ``pole-constrained solution''.

@@

Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$. Clearly, the coordinates of Ceres' pole (\ref{Pole}) can impose a strong constraint on the position angle, as the next solution shows.

@@

At the moment of the occultation, the coordinates (\ref{Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$. Exploration of the parameter space, restricted to ellipses whose position angle lie on this range, results in the pole-constrained solution. The related physical parameters are displayed on the third column of Table~2, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Results of limb fitting to the data of the 2010 and 2013 events.}
  \begin{tabular}{@{}lcccc}
  \hline
     Solution & 2010/Nominal & 2010/Pole-constrained & 2013/Nominal & 2013/Pole-constrained \\
\hline
Equatorial diameter (km) & 982 $\pm$ 14 & 972 $\pm$ 6  & 971 $\pm$ 7  & 971 $\pm$ 7\\
%Equatorial diameter (km) & 491 $\pm$ 7  & 486 $\pm$ 3  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
True oblateness        & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 77 $\pm$ 6    & 78 $\pm$ 6\\
$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 13 $\pm$ 16   & 13 $\pm$ 16\\
$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
\hline
\end{tabular}
\textbf{Notes.} Error bars are at 1-sigma level.
(*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
\end{minipage}
\end{table*}

@@

\section[]{The 2013 occultation}

@@

The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC2 35031360), of magnitude $K = 8.76$. Its ICRF/J2000 position is \citep{Zacharias2004}:
%
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 11^{h}57^{m}52^{s}.7641\\
    \delta = +09\degr 07\arcmin 49\arcsec.835.
  \end{array}
\right.
\end{equation}
%
The occultation could only be visible in the United States, before dawn. The shadow would then cross the Atlantic and Africa, but during daylight time.

\begin{figure}
\includegraphics[scale=0.42]{figures/Ceres_2013.png} 
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA. The big dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right. The blue dots are the sites that have observed the event.\label{Fig: Ceres-2013-map}}
\end{figure}

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for the observing stations of the 2013 event.\label{Tab: obs-2013}}
  \begin{tabular}{@{}lccccc}
  \hline
     Site & Longitude & Telescope: & Camera  & Result & Observer  \\
          & Latitude  & Aperture  & Cycle time & Ingress & \\          
          & Height    & f-ratio    &    S/N    & Egress    & \\          
\hline
 Hampton & 70$\degr$48'59.7'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$53'52.8'' N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 &   \\
            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 &   \\
 & & & & & \\
 Topsfield & 70$\degr$55'16.6'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$37'55.9'' N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 &   \\
            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 &   \\
 & & & & & \\
 Brookline & 71$\degr$08'14.5'' W & 64 cm & Infinity2-1R & Positive & N. Weber \\
  &42$\degr$18'27.4'' N & f/9.6 & 0.015 s     & Duration only &  R. Dantowitz  \\
            & 109 m     &       &     & (14.93 $\pm$ 0.01) s &   \\
 & & & & & \\
 Winchester & 78$\degr$14'39.6'' W & 36 cm & PC164C & Positive & J. Brooks \\
  &39$\degr$16'21.5'' N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 &   \\
            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 &   \\
 & & & & & \\
 Greenbelt & 76$\degr$52'09.4'' W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
  &38$\degr$59'12.1'' N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 & D. Dunham  \\
            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 &   \\
 & & & & & \\
 Alexandria & 77$\degr$02'28.3'' W & 7 cm & Watec120N & Positive & P. Maley \\
  &38$\degr$49'19.1'' N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 &   \\
            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 &   \\
 & & & & & \\
 Owings & 76$\degr$38'06.3'' W & 25 cm & PC164C  & Positive & C. Ellington \\
  &38$\degr$41'26.5'' N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 &   \\
            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 &   \\
 & & & & & \\
 Mechanicsville & 77$\degr$23'06.7'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$41'26.1'' N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 &   \\
            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 &   \\
 & & & & & \\
 Varina & 77$\degr$19'49.3'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$25'58.6'' N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 &   \\
            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 &   \\

@@

All videos were converted to FITS images and the photometry of the target was obtained via the PRAIA \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of a reference star when such was available on the field.
